---
title: "heart disease prediction "
author: "Precious Ebu"
date: "2/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#loading required libraries
```{r}
library(tidyverse)
library(readxl)
library(psych)
library(ggpubr)
library(lattice)
library(ggplot2)
library(caret)
```
#importing the dataset
```{r}
heart <- read.csv("C:\\Users\\user\\Documents\\DATALAB\\Datasets\\New Dataset\\heart_disease_health_indicators_BRFSS2015.csv")
```

#first glimpse of the dataset
```{r}
head(heart)
```

#viewing the data
```{r}
view(heart)
```

#getting insights
```{r}
str(heart)
```
#getting further insights
```{r}
glimpse(heart)
```
#checking the decriptive statistics
```{r}
summary(heart)
```
#checking for missing values(NA)
```{r}
which(is.na(heart))
```
#checking the sum of NAs
```{r}
sum(is.na(heart))
```
#subsetting the numeric variables
```{r}
heart[c(5,16,17,20)]
```
#checking for unique values
```{r}
unique(heart$GenHlth)
```

#converting numeric columns to integer datatype
```{r}
heart[c(5,16,17,20)] <- lapply(heart[c(5,16,17,20)], as.integer)
```

#confirming the conversion
```{r}
str(heart)
```
#checking unique values
```{r}
unique(heart$Education)
```

```{r}
unique(heart$Income)
```

#converting categorical variables to factor
```{r}
heart[c(1:4,6:15,18,19)] <- lapply(heart[c(1:4,6:15,18,19,21,22)], as.factor)
```

#confirming the conversion
```{r}
glimpse(heart)
```
###EXPLORATORY DATA ANALYSIS (EDA)
#saving the numeric columns into a variable name
```{r}
heart_cont <- heart[c(5,16,17,20)]
```

#visualizing outliers
```{r}
boxplot(heart_cont)
```

#saving outliers in BMI into a variable
```{r}
outliers <- boxplot(heart$BMI, plot = FALSE)$out 
```

#converting outliers to NAs
```{r}
heart$BMI[heart$BMI %in% outliers] <- NA
```

#checking sum of NAs after the conversion
```{r}
sum(is.na(heart$BMI))
```

#saving outliers in "MentHlth" into a variable
```{r}
outliers2 <- boxplot(heart$MentHlth, plot = FALSE)$out 
```

#converting outliers to NAs
```{r}
heart$MentHlth[heart$MentHlth %in% outliers2] <- NA
```

#checking the sum of NAs after conversion
```{r}
sum(is.na(heart$MentHlth))
```

#saving outliers into a variable
```{r}
outliers3 <- boxplot(heart$PhysHlth, plot = FALSE)$out 
```

#converting outliers to NAs
```{r}
heart$PhysHlth[heart$PhysHlth %in% outliers3] <- NA
```

#checking the sum of NAs after the converion
```{r}
sum(is.na(heart$PhysHlth))
```

#replacing NAs with the column mean
```{r}
heart$BMI[is.na(heart$BMI)] <-  mean(heart$BMI, na.rm = TRUE)
heart$MentHlth[is.na(heart$MentHlth)] <- mean(heart$MentHlth, na.rm = TRUE)
heart$PhysHlth[is.na(heart$PhysHlth)] <- mean(heart$PhysHlth, na.rm = TRUE)
```

#confirming there are no outliers 
```{r}
heart[which(heart$BMI %in% outliers),]
heart[which(heart$MentHlth %in% outliers2),]
heart[which(heart$PhysHlth %in% outliers3),]
```

#saving the numeric columns into a variable name
```{r}
heart_num <- heart[c(5,16,17,20)]
```

#visualizing outliers after replacement
```{r}
boxplot(heart_num)
```

#checking the descriptive statistics after replacement
```{r}
describe(heart_num)
```

#confirming there are no NAs
```{r}
which(is.na(heart))
```

#DATA VISUALIZATION
#plotting all variables against the target variable "HeartDiseaseorAttack"
```{r}
ggplot(heart, aes(x = HeartDiseaseorAttack))+ 
  geom_bar(mapping = aes(fill=HeartDiseaseorAttack)) + geom_text(aes(y=..count..-200, 
                label = paste0(round(prop.table(..count..),4) * 100, '%')), 
            stat = 'count', position = position_dodge(.1), size =3)
```

#saving variables into a name and plotting bar chart of all variables
```{r}
variables <-list("HighBP", "HighChol", "CholCheck", "Smoker", "Stroke", "Diabetes", "PhysActivity", "Fruits", "Veggies", "HvyAlcoholConsump", "AnyHealthcare", "NoDocbcCost", "GenHlth", "DiffWalk", "Sex", "Education", "Income") 
plot <- list()

 for (i in variables){
 plot <-  ggplot(data = heart, aes_string(x=i, fill = as.factor(heart$HeartDiseaseorAttack)))+
  geom_bar(position = "stack") + 
  geom_text(aes(y=..count..-200, 
                label = paste0(round(prop.table(..count..),4) * 100, '%')), 
            stat = 'count', position = position_dodge(.1), size =3)
   print(plot)
}
```

#checking for relationships between categorical variables
```{r}
chisq.test(heart$CholCheck, heart$HeartDiseaseorAttack) 
```

```{r}
chisq.test(heart$Smoker, heart$HeartDiseaseorAttack)

```

```{r}
chisq.test(heart$Stroke, heart$HeartDiseaseorAttack)

```

```{r}
chisq.test(heart$Diabetes, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$HighBP, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$HighChol, heart$HeartDiseaseorAttack)
```

```{r}
colnames(heart)
```

```{r}
chisq.test(heart$PhysActivity, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$Fruits, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$Veggies, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$HvyAlcoholConsump, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$AnyHealthcare, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$NoDocbcCost, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$GenHlth, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$DiffWalk, heart$HeartDiseaseorAttack)
```


```{r}
chisq.test(heart$Sex, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$Education, heart$HeartDiseaseorAttack)
```

```{r}
chisq.test(heart$Income, heart$HeartDiseaseorAttack)
```

#checking for correlation between numerical variables
```{r}
cor(heart$BMI, heart$MentHlth)
```

```{r}
cor(heart$MentHlth, heart$PhysHlth)
```

```{r}
cor(heart$BMI, heart$PhysHlth)
```

```{r}
cor(heart$BMI, heart$Age)
```

```{r}
cor(heart$MentHlth, heart$Age)
```

```{r}
cor(heart$PhysHlth, heart$Age)
```

```{r}
heart_cor <- cor(heart_num)
```

#loading a library
```{r}
library(corrplot)
```

#plotting a correlation matrix
```{r}
corrplot::cor.mtest(heart_num)
```

#visualizing the matrix
```{r}
corrplot(heart_cor)
```

#carrying out a one-way ANOVA
```{r}
result <- aov(BMI~ HeartDiseaseorAttack, data = heart)
summary(result)
```

#plotting a heat map
```{r}
heatmap(heart_cor)
```

#tabulating the target variable
```{r}
table(heart$HeartDiseaseorAttack)
```

#plotting the target variable on a bar chart
```{r}
ggplot(data = heart, aes(fill = HeartDiseaseorAttack)) +
    geom_bar(aes(x = HeartDiseaseorAttack))+
    ggtitle("Number of samples in each class", subtitle = "Heart Disease data")+
    xlab("")+
    ylab("Samples")+
    scale_y_continuous(expand = c(0,0))+
    scale_x_discrete(expand = c(0,0))+
    theme(legend.position = "none", 
         legend.title = element_blank(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         panel.background = element_blank())
```
#DATA MODELLING AND MACHINE LEARNING
#loading a library for splitting
```{r}
library(caTools)
```

```{r}
#making this example reproducible
set.seed(265)

#Using 70% of dataset as training set and remaining 30% as testing set
sample <- sample.split(heart$HeartDiseaseorAttack, SplitRatio = 0.7)
train_data <- subset(heart, sample == TRUE)
test_data  <- subset(heart, sample == FALSE)
```

#checking the dimension of the train and test data
```{r}
dim(train_data)
dim(test_data)
```

#tabulating the number of cases of the target variable in both sets
```{r}
prop.table(table(train_data$HeartDiseaseorAttack))
prop.table(table(test_data$HeartDiseaseorAttack))
```
#FEATURE SELECTION
```{r}
library(Boruta)
```

```{r}
train <- train_data[,c(-2,-3,-6)]
test <- test_data[,c(-2,-3,-6)]
```

```{r}
dim(train)
dim(test)
```

#CLASS BALANCING
#balancing the imbalanced class using ROSE
```{r}
library(ROSE)
```

#fitting the data into the rose model
```{r}
heart_rose <- ROSE(HeartDiseaseorAttack~., data = train, seed=123)$data
table(heart_rose$HeartDiseaseorAttack)
```

#visualizing the balance
```{r}
ggplot(data = heart_rose, aes(fill = HeartDiseaseorAttack))+
    geom_bar(aes(x = HeartDiseaseorAttack))+
    ggtitle("Number of samples in each class after ROSE", subtitle="Total samples:  177576")+
    xlab("")+
    ylab("Samples")+
    scale_y_continuous(expand = c(0,0))+
    scale_x_discrete(expand = c(0,0))+
    theme(legend.position = "none", 
          legend.title = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank())
```

```{r}
library(randomForest)
```

#fitting the balanced data into random forest algorithm
```{r}
rfrose <- randomForest(HeartDiseaseorAttack~., data=heart_rose)
```

#getting the confusion matrix
```{r}
caret::confusionMatrix(predict(rfrose, test), test$HeartDiseaseorAttack, positive = '1')
```

#Building Logistic Regression Models
```{r}
rose_logmodel <- glm(HeartDiseaseorAttack~., data = heart_rose, family = binomial(link = "logit"))
```

#getting the model summary
```{r}
summary(rose_logmodel)
```

#running anova on the model
```{r}
anova(rose_logmodel, test = "Chisq")
```

#using the LR model to predict
```{r}
pred_results <- predict(rose_logmodel, test, "link")

pred_results <- ifelse(pred_results > 0.5,1,0)
```

#getting the model accuracy
```{r}
misClasificError <- mean(pred_results != test$HeartDiseaseorAttack)
print(paste('Accuracy',1-misClasificError))
```

#viewing the top rows of the predicted
```{r}
head(pred_results)
```

#converting the predicted to factor
```{r}
pred_results <- as.factor(pred_results)

str(pred_results)
```

#getting the confusion matrix
```{r}
caret::confusionMatrix(pred_results, test$HeartDiseaseorAttack, positive="1", mode="everything")
```

#loading a library for cross validation
```{r}
library(e1071) 
```

#fitting the data with a repeated CV model
```{r}
rose_cv <- train(HeartDiseaseorAttack~., data = heart_rose, method = "glm", family = "binomial", metric="Accuracy", trControl = trainControl(method = "repeatedcv", number = 10, repeats = 5)) 

rose_cv
```

#saving the prediction into a variable name
```{r}
pred_cv <- predict(rose_cv, test)
```

#converting the predicted to factor
```{r}
pred_cv <- as.factor(pred_cv)
str(pred_cv)
```

#getting the confusion matrix 
```{r}
caret::confusionMatrix(pred_cv, test$HeartDiseaseorAttack, positive="1",mode="everything")
```

```{r}

```


